---
title: "Final"
output: html_document
date: "2024-12-13"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Project 1
```{r}
  # Reading in files
ppg <- read.csv('ppg2008.csv')
rownames(ppg) <- ppg[,1]
ppg<- ppg[,-1]
ppg2<- read.csv('ppg2008.csv')
```

```{r}
  # Installing/calling necessary packages
install.packages('Rtsne')
library(Rtsne)
library(ggplot2)
```

```{r}
  # Performing t-SNE dimension reduction
ppg_tsne <- Rtsne(ppg, perplexity = 16)

  # Plotting values resulting from dimension reduction
tsne_df <- data.frame(
  X = ppg_tsne$Y[, 1],
  Y = ppg_tsne$Y[, 2]
)
ggplot(tsne_df, aes(X,Y))+
  geom_point()

  # Instaling packages necessary for PCA 
install.packages("factoextra")
library(factoextra)

ppg_pca<- princomp(ppg)
  # Getting an idea of the PCA results by looking and loadings and visualizing it
summary(ppg_pca)
ppg_pca$loadings[,1:4]
fviz_pca_var(ppg_pca)

```

  Based on these PCA results, the top 4 components are explained most by number of games, portion of team 3 pt attempts the player contributed, total rebounds and field goal attempts. These components encapsulate 95 percent of the data, so they will be analyzed further.

```{r}
  # Isolating the desired stats and cleaning up data presentation
pca_iso_col <- as.matrix(cbind(ppg$G,ppg$X3PA,ppg$TRB,ppg$FGA))
rownames(pca_iso_col)<- rownames(ppg)
colnames(pca_iso_col) <- c('G','X3PA','TRB','FGA')
cor(pca_iso_col[,1:4],pca_iso_col[,1:4], method = 'pearson')

  # Performing k-means clustering on the isolated data
  # First, the number of clusters need to be determined by how much the within cluster sum of squares is reduced with k clusters
k.max <- 15
wss <- sapply(1:k.max, 
              function(k){kmeans(pca_iso_col, k, nstart=50,iter.max = 15)$tot.withinss})
plot(1:k.max, wss,
     type="b", pch = 19, frame = FALSE, 
     xlab="Number of clusters K",
     ylab="Total within-clusters sum of squares")

  # installing/calling packages for color scheme
install.packages("wesanderson")
library(wesanderson)

  # Performing k-means and plotting the clusters of each variable against the others
km_clust<- kmeans(pca_iso_col[,1:4],3)

ggplot(data = pca_iso_col, aes(x = G, y = X3PA, color = as.factor(km_clust$cluster))) +
  geom_point(size = 3)+
  scale_color_manual(values = wes_palette("Moonrise3",5,type = "discrete"),name = "Player Cluster") +
  labs(title = "Player Grouping Based on Games and X3PA ",
       x = "Number of Games Played",
       y = "% Contribution To Total Team 3 Point Attempts")+
  geom_text(size = 3, label= rownames(ppg), check_overlap = TRUE, nudge_x = 0.5, nudge_y = -0.5)

ggplot(data = pca_iso_col, aes(x = G, y = TRB, color = as.factor(km_clust$cluster))) +
  geom_point(size = 3)+
  scale_color_manual(values = wes_palette("Moonrise3",5,type = "discrete"),name = "Player Cluster") +
  labs(title = "Player Grouping Based on Games and TRB ",
       x = "Number of Games Played",
       y = 'Total Rebounds')+
  geom_text(size = 3, label= rownames(ppg), check_overlap = TRUE, nudge_x = 0.5, nudge_y = -0.5)


ggplot(data = pca_iso_col, aes(x = G, y = FGA, color = as.factor(km_clust$cluster))) +
  geom_point(size = 3)+
  scale_color_manual(values = wes_palette("Moonrise3",5,type = "discrete"),name = "Player Cluster") +
  labs(title = "Player Grouping Based on Games and FGA ",
       x = "Number of Games Played",
       y = 'Field Goal Attempts')+
  geom_text(size = 3, label= rownames(ppg), check_overlap = TRUE, nudge_x = 0.5, nudge_y = -0.5)

ggplot(data = pca_iso_col, aes(x = X3PA, y = TRB, color = as.factor(km_clust$cluster))) +
  geom_point(size = 3)+
  scale_color_manual(values = wes_palette("Moonrise3",5,type = "discrete"),name = "Player Cluster") +
  labs(title = "Player Grouping Based on X3PA and TRB ",
       x = "% Contribution To Total Team 3 Point Attempts",
       y = 'Total Rebounds')+
  geom_text(size = 3, label= rownames(ppg), check_overlap = TRUE, nudge_x = 0.5, nudge_y = -0.5)


ggplot(data = pca_iso_col, aes(x = X3PA, y = FGA, color = as.factor(km_clust$cluster))) +
  geom_point(size = 3)+
  scale_color_manual(values = wes_palette("Moonrise3",5,type = "discrete"),name = "Player Cluster") +
  labs(title = "Player Grouping Based on X3PA and FGA ",
       x = "% Contribution to Team 3 Point Attempts",
       y = 'Field Goal Attempts')+
  geom_text(size = 3, label= rownames(ppg), check_overlap = TRUE, nudge_x = 0.5, nudge_y = -0.5)


ggplot(data = pca_iso_col, aes(x = TRB, y = FGA, color = as.factor(km_clust$cluster))) +
  geom_point(size = 3)+
  scale_color_manual(values = wes_palette("Moonrise3",5,type = "discrete"),name = "Player Cluster") +
  labs(title = "Player Grouping Based on TRB and FGA ",
       x = "Total Rebound",
       y = 'Field Goal Attempts')+
  geom_text(size = 3, label= rownames(ppg), check_overlap = TRUE, nudge_x = 0.5, nudge_y = -0.5)

```
  Based on these findings, it is fair to say that using this data set as is for unbiased analysis of these players is unsuitable. Firstly, the t-SNE reduction produced a set of points to define each of the 50 players, but upon plotting them, there is no pattern to be discerned. And with multiple runs of the program, the variety of points/plots did not produce any more information than the first one. Because of this, a PCA reduction was performed, which revealed that more than 95% of the variation in the data could be captured by the first four components. These components were explained most by Games Played, % Contribution To Total Team 3 Point Attempts, Total Rebound, and Field Goal Attempts, respectively. The data was then clustered using k-means. The cluster number was determined using the "elbow" method, which showed the optimal number of clusters may be 2, but I opted for 3 to be sure that any patterns were revealed. However, the k-means clustering did not reveal any relationship between these variables either. The first three graphs have clear clusters, but there is no correlation between them. The last three graphs do not have sensible clustering, as some points that belong to one cluster are somehow in the midst of another cluster, and overall there is no pattern or correlation between the points. Additionally, this data set is not a good candidate for GLMs either, as the response variable would be a name. Even when being coerced by using as.factor(), it is not possible. I believe that this data is lacking an additional factor, which is related to each players playing style or their position, so that they're grouped already in that manner and subsequent analysis can be performed then. Or perhaps, this is a data set where context in which it was collected needs to be completely understood by the analyzer. Some other factors that seem important to consider are: the market (team/location) from which the players come from, players' body metrics, injury, and the fact that the way someone plays a sport is unique so a catch-all comparison between players will almost always neglect a few statistics or niche knowledge is needed to pick the right ones for analysis. 


#Project 3
```{r}
  # Calling necessary packages
install.packages("BiocManager")
BiocManager::install("preprocessCore")
```

##1.
```{r}
  # Reading in data
mrna <- read.csv("Mnemiopsis_count_data.csv")
rownames(mrna) <- mrna[,1]
mrna <- mrna[,-1]
mrna1<- mrna

  # Removing low expressing genes based on an average expression value of =< 1
mrna1$averages <- apply(mrna1, 1,function(x)mean(x))
mrna1<- mrna1[mrna1$averages >= 1,]

  # Normalizing data using quantiles and checking that all column averages are the same
library(preprocessCore)
normalized_mrna1<- normalize.quantiles(as.matrix(mrna1))
colnames(normalized_mrna1) <- colnames(mrna1)
rownames(normalized_mrna1) <- rownames(mrna1)
normalized_mrna1 <- normalized_mrna1[,-9]     # removing column that held averages, as it is no longer necessary
colMeans(normalized_mrna1)

  # Creating distance trees using Spearman and Euclidean Distance

pcor_NormMrna1 <- as.dist(1-cor(t(normalized_mrna1),method = "spearman")/2)
pcor_NormMrna1_clst <- hclust(pcor_NormMrna1,method = "complete")
plot(pcor_NormMrna1_clst, cex = 0.3, ylab ="Spearman Correlation Distance", main = 'Mnemiopsis Gene Expression Values', xlab = "Genes", sub = NA)


distanced<- dist(normalized_mrna1)
ppg_clst <- hclust(distanced,method ='complete')
plot(ppg_clst, hang = -1, cex = 0.3, ylab ="Euclildean Distance", main = 'Mnemiopsis Gene Expression', xlab = "Genes", sub = NA)

pcor_NormMrna2 <- as.dist(1-cor(normalized_mrna1, method = "spearman")/2)
pcor_NormMrna2_clst <- hclust(pcor_NormMrna2, method = "complete")
plot(pcor_NormMrna2_clst, cex = 1, ylab ="Spearman Correlation Distance", main = 'Mnemiopsis Gene Expression', xlab = "Gene Group", sub = NA)

distanced<- dist(t(normalized_mrna1))
ppg_clst <- hclust(distanced,method ='complete')
plot(ppg_clst, hang = -1, cex = 1, ylab ="Euclildean Distance", main = 'Mnemiopsis Gene Expression', xlab = "Gene Group", sub = NA)

```
##2.
```{r}
  # Installing necessary packages for heat map
install.packages("pheatmap")
```
```{r}
# Calling necessary packages and creating heat map
library(pheatmap)
library(RColorBrewer)
pheatmap(normalized_mrna1, col = colorRampPalette(brewer.pal(n= 9, name = 'BuPu'))(256), scale = "row", fontsize_row = 1, cluster_cols = pcor_NormMrna2_clst,cluster_rows = pcor_NormMrna1_clst, legend_breaks =c(-2,0,2), legend_labels = c("lower","medium","high"), main = 'Aboral and Oral Gene Expression Levels' )

```
##3.
###a.
```{r}
  # Installing packages necessary
BiocManager::install("DESeq2")
```
```{r} 
  # Calling necessary packages and preparing RNA count data for analysis
library(DESeq2)
mrna2 <- read.csv("Mnemiopsis_count_data.csv")
rownames(mrna2) <- mrna2[,1]
mrna2<- mrna2[,-1]

colsdata <- read.csv("Mnemiopsis_col_data.csv")
rownames(colsdata) <- colsdata[,1]
rownames(colsdata) <- sub('-','',rownames(colsdata))
colsdata<- colsdata[,-1]
colnames(colsdata)[2]<- "group"      # condition changed to group for compatibility with Glimma 

  # Creating DESeq data set
des_mrna<- DESeqDataSetFromMatrix(countData = mrna2,
                                  colData = colsdata,
                                  design=~group)

  # Isolating the genes with average expression >=1
keptcounts <- rowMeans(counts(des_mrna))>=1
des_mrna<- des_mrna[keptcounts,]

  # Running DESeq analysis, storing results, and creating a results data frame
des_mrna <- DESeq(des_mrna)
rez<- results(des_mrna, alpha = 0.05)
rez_df <- as.data.frame(rez)

  # Subsetting the log fold change adjusted p-value, and base mean columns for further manipulation
lfc_padj_subset <- data.frame(l2fc = rez_df$log2FoldChange,
                            padj = rez_df$padj,
                            basemean = rez_df$baseMean)
rownames(lfc_padj_subset)<-rownames(rez_df)
most_changing <- lfc_padj_subset[order(lfc_padj_subset$l2fc, decreasing = TRUE),]


  # Listing the names of the genes that are the "most-changing", meaning that they had the greatest log fold change
rownames(head(most_changing[,'l2fc',drop=FALSE],10))
```
###b.
```{r}
  # Identifying the top 5 housekeeping genes. This was done by finding the midpoint of the log fold change column (as the data is already ordered) and taking the ten genes above and below that. This represents genes that are continuously expressed as the log fold change in counts is close to zero. The names of the gene highest counts were pulled from an ordered basemean column, and is reported.
mid<- length(most_changing$l2fc)/2
hk_genes <- rbind(most_changing[mid+c(1:10),],most_changing[mid-c(1:10),])
hk_genes <- hk_genes[order(hk_genes$basemean,decreasing = TRUE),]
rownames(head(hk_genes[,'basemean', drop=FALSE]),5)
```
  
```{r message=FALSE}
  # Installing necessary packages for DESeq visualization
BiocManager::install("Glimma")
```
```{r}
  # Calling necessary packages for DESeq visualization
library(Glimma)
glimmaMA(des_mrna)
```

###c/d/e.
  The findings made in the midterm and with this DESeq analysis are perfectly consistent. The top 5 highest expressing genes in the midterm are corroborated by the Glimma visualization of the DESeq data by sorting the logCPM column. In the midterm the correlation between condition groups were discussed, where it was determined that all groups are similar but there is more variation between groups than within. This is confirmed by the trees, where the Euclidean and Spearman correlation distances show greater distance between the nodes that connect the two different groups and the nodes that connect the variables within the aboral and oral groups. However, the trees reveal that in terms of Euclidean distance, the oral group is all the same for the most part, while in the aboral group, aboral1 is noticeably different from the rest of the aboral group. This is interesting and warrants investigation into why that is. A tree of the genes themselves, as seen above, is too cluttered to meaningfully discuss any correlations between genes, but in both one can observe a slight separation of the genes into groups, which imply that some genes are related to processes or pathways related to the individual aboral/oral conditions. In the midterm, we were tasked to find the most and least variable genes. Unsurprisingly, in the Glimma plot, the least variable genes are all in the center of the plot, where the fold change is 0 and the logCPM is between 4-6. This shows they truly don't vary and are likely the group of housekeeping genes. The most variable genes found in the midterm, however, are all downregulated in the Glimma plot, with a high logCPM, save for ML174731a which was neither up or down regulated. This would definitely necessitate some type of investigation, as it seems like these samples may have been taken from developing Mnemiopsis embryos based on the housekeeping genes. Personally, I would run more experiments and do further research on this set of more variable genes, as this finding was very surprising.
